# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data
The data are stored in zip archive, which contains only one file, using read.csv() 
and unz() connection to load data into variable activity.

```{r}
activity <- read.csv(unz("activity.zip","activity.csv"))
```

Then apply str() to show the data.
```{r}
str(activity)
```

Obviously, activity$date should not be a factor, convert it to POSIX date-time objects with strptime(). 
And check it again.
```{r}
activity$date <- as.POSIXct(strptime(activity$date,format="%Y-%m-%d"))
str(activity)
```

## What is mean total number of steps taken per day?
Using aggregate() to compute the mean of total number steps taken per day, parameter na.rm set to TRUE
to ignore missing values.

```{r}
stepsdate <-
  aggregate(formula = steps~date, data = activity,
            FUN = sum, na.rm=TRUE)
```

Then, make a histogram.
```{r}
barplot(stepsdate$steps,
        main="Total Number of Steps Per Day", 
        xlab="Steps Per Day")
```
Mean and median of the number of steps take per day are following:
```{r}
mean(stepsdate$steps)
median(stepsdate$steps)
```

## What is the average daily activity pattern?
Aggregate the steps per interval, calculating the mean across the days
```{r}
stepsinterval <- 
  aggregate(formula=steps~interval, data=activity,
            FUN=mean, na.rm=TRUE)
```
Make a time series plot using stepsinterval.
```{r}
plot(stepsinterval$interval, stepsinterval$steps, 
     type = "l", xlab = "5-Minute Intervals",  ylab = "Steps")
```
The 5-min interval that contains, on average, the maximum number of steps is:
```{r}
stepsinterval[which(stepsinterval$steps==max(stepsinterval$steps)),]
```

## Imputing missing values
In previous analysis, records with missing values are removed by na.rm=TRUE.
```{r}
rownum.na<-sum(is.na(activity))
rownum<-nrow(activity)
```
Total number of rows with NAs is `r rownum.na`. And total row number is `r rownum`. 

Creating new data set by replacing NAs with mean value for respective interval:
```{r}
activity_imputed <- activity

for (i in seq(1:dim(activity_imputed)[1])) {
    if (is.na(activity_imputed[i, ]$steps)) {
        activity_imputed[i, ]$steps <- stepsinterval[stepsinterval$interval == 
            activity_imputed[i, ]$interval, ]$steps
    }
}
```
Aggregating daily activity from the imputed dataset and plot it.
```{r}
stepsdateimputed <-
  aggregate(formula = steps~date, data = activity_imputed,
            FUN = sum)
barplot(stepsdateimputed$steps,
        main="Total Number of Steps Per Day", 
        xlab="Steps Per Day")
```
Compare with dataset which NA records removed, mean and median are:
```{r}
mean(stepsdateimputed$steps)
median(stepsdateimputed$steps)
```
Mean remain the same and median increased by 1.

## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable in the dataset with two levels, then divide the dataset into 2 subset.
Call Sys.setlocale() since my system is Chinese.
```{r}
Sys.setlocale("LC_TIME", "English")
dindicator <- weekdays(as.Date(activity_imputed$date, "%Y-%m-%d"))
dindicator[dindicator == "Sunday" | dindicator == "Saturday"] <- "weekend"
dindicator[dindicator != "weekend"] <- "weekday"
activity_imputed$weekday_type <- as.factor(dindicator)

activity_imputed_weekday <- activity_imputed[activity_imputed$weekday_type == "weekday", ]
interval_activity_weekday <- aggregate(steps ~ interval, data = activity_imputed_weekday, mean)
activity_imputed_weekend <- activity_imputed[activity_imputed$weekday_type == "weekend", ]
interval_activity_weekend <- aggregate(steps ~ interval, data = activity_imputed_weekend, mean)
```
Plotting it.
```{r}
par(mfrow = c(2, 1))
plot(x = interval_activity_weekday$interval, y = interval_activity_weekday$steps, 
    type = "l", xlab = "Weekday intervals", ylab = "Number of Steps")    
plot(x = interval_activity_weekend$interval, y = interval_activity_weekend$steps, 
    type = "l", xlab = "Weekend intervals", ylab = "Number of Steps")
```

Thanks for your time
