# Consequences of Storm Events

## Synopsis

This analysis try to answer two questions:

1. Across the United States, which types of events (as indicated in the EVTYPE 
variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic 
consequences?

After exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) 
storm database, it shows that of all severe weather events, HEAT and TORNADO. are the most 
harmful to population health. But to the greatest economic 
consequences, HURRICANE and HEAT are the most damage types.

## Data Processing
Read data by method read.csv() after download the bz2 file and wrap it by tbl_df().
 
```{r cache=TRUE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(gridExtra)
Sys.setlocale("LC_TIME", "English")
stormdata <- read.csv("repdata-data-StormData.csv.bz2", header=T, stringsAsFactors=F)
stormdata <- tbl_df(stormdata)
names(stormdata)
stormdatancol<-ncol(stormdata);stormdatancol
nrow(stormdata)
```
This database contains 37 observations of major weather events in the United States, 
including when and where they occured, as well as estimates of any fatalities, injuries, and property
damage.

According to the description, data was not gathered on all of the categories until 1996. So to drop events 
prior to 1996, and variables to do the analysis also selected.
```{r cache=TRUE}
stormdata1996 <- select(stormdata, BGN_DATE, EVTYPE, FATALITIES:CROPDMGEXP)
stormdata1996$BGN_DATE <- mdy_hms(stormdata1996$BGN_DATE)
stormdata1996 <- filter(stormdata1996, year(BGN_DATE) >= 1996)
stormdata1996 <- filter(stormdata1996, substr(EVTYPE,1,7) != "Summary")
names(stormdata1996)
ncol(stormdata1996)
nrow(stormdata1996)
```

The EVTYPE contains inconsistent data, following code reassign event type to fields EVLABEL.
```{r cache=TRUE}
stormdata1996$EVTYPE <- toupper(stormdata1996$EVTYPE)
stormdata1996$EVLABEL <- "OTHER"
stormdata1996$EVLABEL[grepl("FIRE",stormdata1996$EVTYPE)] <- "FIRE"
stormdata1996$EVLABEL[grepl("DUST",stormdata1996$EVTYPE)] <- "DUST"
stormdata1996$EVLABEL[grepl("FLOOD|SURGE",stormdata1996$EVTYPE)] <- "FLOOD"
stormdata1996$EVLABEL[grepl("HAIL",stormdata1996$EVTYPE)] <- "HAIL"
stormdata1996$EVLABEL[grepl("WIND",stormdata1996$EVTYPE)] <- "HIGH WIND"
stormdata1996$EVLABEL[grepl("WARM|HOT|DROUGHT|DRY|DRI|HEAT|HIGH",
                      stormdata1996$EVTYPE)] <- "HEAT"
stormdata1996$EVLABEL[grepl("COLD|FROST",stormdata1996$EVTYPE)] <- "COLD"
stormdata1996$EVLABEL[grepl("SNOW|BLIZZARD|WINT*MIX|WINT*STOR|ICE",
                      stormdata1996$EVTYPE)] <- "SNOWSTORM"
stormdata1996$EVLABEL[grepl("TORN|SPOUT|WALL|FUNNEL",
                      stormdata1996$EVTYPE)]<-"TORNADO"
stormdata1996$EVLABEL[grepl("HURRICANE|TYPHOON|TROP",
                      stormdata1996$EVTYPE)]<-"HURRICANE"
stormdata1996$EVLABEL[grepl("THUN|TSTM|MICROB|RAIN",
                      stormdata1996$EVTYPE)] <- "THUNDERSTORM"
stormdata1996$EVLABEL[grepl("WINTER STORM",
                      stormdata1996$EVTYPE)] <- "WINTER STORM"
stormdata1996$EVLABEL[grepl("LIGHTNING",
                      stormdata1996$EVTYPE)] <- "LIGHTNING"
stormdata1996$EVLABEL[grepl("WINTER WEATHER|WINTER MIX|WINTRY MIX|WINTERY MIX",
                      stormdata1996$EVTYPE)] <- "WINTER WEATHER"
stormdata1996$EVLABEL[grepl("WET",
                      stormdata1996$EVTYPE)] <- "WET"
stormdata1996$EVLABEL[grepl("FOG",
                      stormdata1996$EVTYPE)] <- "FOG"
stormdata1996$EVLABEL[grepl("RIP CURRENT",
                      stormdata1996$EVTYPE)] <- "RIP CURRENT"
stormdata1996$EVLABEL[grepl("STREAM FLD|STRM FLDG",
                      stormdata1996$EVTYPE)] <- "STREAM FLD"
table(stormdata1996$EVLABEL)
```

Two columns PROPDMG(property damage) and CROPDMG(crop damage) should take account of their multiplier in PROPDMGEXP and CROPDMGEXP.
```{r cache=TRUE}
table(stormdata1996$PROPDMGEXP)
table(stormdata1996$CROPDMGEXP)

multiplier <- function (x, y) { 
  x <- as.numeric(x)
  if (y == "K") {
    x <- x*10^3
  } else if (y == "M") {
    x <- x*10^6
  } else if (y == "B") {
    x <- x*10^9
  } else {
    x
  }
}

stormdata1996$PROPDMGRESULT <- mapply(multiplier, stormdata1996$PROPDMG, stormdata1996$PROPDMGEXP)
stormdata1996$CORPDMGRESULT <- mapply(multiplier, stormdata1996$CROPDMG, stormdata1996$CROPDMGEXP)
```

FATALITIES and INJURIES should be convert to numeric.
```{r cache=TRUE}
stormdata1996$FATALITIESNUM <- as.numeric(stormdata1996$FATALITIES)
stormdata1996$INJURIESNUM <- as.numeric(stormdata1996$INJURIES)
```


## Results
Now we can answer the two questions. For the first one, FATALITIES and INJURIES data are aggregated by EVLABEL.
```{r cache=TRUE}
stormdatahealth <- aggregate(cbind(FATALITIESNUM, INJURIESNUM) ~ EVLABEL, data=stormdata1996, sum)
stormdatahealth <- filter(stormdatahealth, (FATALITIESNUM > 0 | INJURIESNUM > 0) )
```

Then plot it.

```{r cache=TRUE}
g1 <- ggplot(arrange(stormdatahealth, desc(FATALITIESNUM))[1:10,], aes(x=reorder(EVLABEL, desc(FATALITIESNUM)), y=FATALITIESNUM)) + 
        geom_bar(fill="indianred", colour="black", stat="identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
        xlab("Event Type") +
        ylab("Fatalities") 

g2 <- ggplot(arrange(stormdatahealth, desc(INJURIESNUM))[1:10,], aes(x=reorder(EVLABEL, desc(INJURIESNUM)), y=INJURIESNUM)) + 
        geom_bar(fill="steelblue3", colour="black", stat="identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust=1)) +
        xlab("Event Type") + 
        ylab("Injuries") 

grid.arrange(g1, g2, ncol=2, main="Fatalities and Injuries by Event Type")
```
The figure shows that the most harmful events are HEAT and TORNADO.

For the last questions, harmful event to economics is illustrated by same method. HURRICANE is the most damage type for property,
and HEAT, for corp damage. 
```{r cache=TRUE}
stormdataeconomics <- aggregate(cbind(PROPDMGRESULT, CORPDMGRESULT) ~ EVLABEL, data=stormdata1996, sum)
stormdataeconomics <- filter(stormdataeconomics, (PROPDMGRESULT > 0 | CORPDMGRESULT > 0) )

g1 <- ggplot(arrange(stormdataeconomics, desc(PROPDMGRESULT))[1:10,], aes(x=reorder(EVLABEL, desc(PROPDMGRESULT)), y=PROPDMGRESULT)) + 
        geom_bar(fill="indianred", colour="black", stat="identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
        xlab("Event Type") +
        ylab("Property") 

g2 <- ggplot(arrange(stormdataeconomics, desc(CORPDMGRESULT))[1:10,], aes(x=reorder(EVLABEL, desc(CORPDMGRESULT)), y=CORPDMGRESULT)) + 
        geom_bar(fill="steelblue3", colour="black", stat="identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust=1)) +
        xlab("Event Type") + 
        ylab("Crop") 

grid.arrange(g1, g2, ncol=2, main="Economic Damage by Event Type")
```
